\documentclass[10pt,a4paper]{article}

\usepackage[english]{babel}


\begin{document}

\begin{center}
%\title{ \bf Granular Materials }
%\author{ Pawel Gniewek }
%\maketitle
\textsc{\Large Granular Materials - Biofilms}\\[1.0cm]
\textsc{\LARGE Pawel Gniewek}\\[1.0cm]
\vspace{0.5cm}
\today

\newpage
\tableofcontents
\end{center}

\newpage

\section{Motivation}
\subsection{General information}
\begin{itemize}
 \item Bacteria (and biofils) are everywhere: Dental plaque, hydrothermal vents, tumors
 \item $10^7 - 10^9$ species
 \item Total biomass: $10^{11}$ tons, and $\sim 10^{30}$ cells.
 \item 10x more cells in human body than human cells
\end{itemize}

\subsection{Modeling}
\begin{itemize}
 \item ODE/PDE mass transport and biochemical reactions models
 \begin{itemize}
  \item reaction/diffusion for biomass and substrates
  \item fluid dynamics for the liquid flow
  \item structural mechanics for biofilm growth and EPS
 \end{itemize}
 \item Discrete-element models
  \begin{itemize}
  \item Cellular Automata (rule based)
  \item individual cells
  \item biomass "particles"
 \end{itemize}
 \item Similarity to Granular materials: nails, medicine capsules, colloids and nono-particles, microtubules
 \item Modeling by MD-kind simulations - including size dependent friction.
 \item Cells grow(elongate) exponentailly $ l = l_0 e^{\alpha t}$, and divide when $l > l_m \approx 2l_0$.
 \item A bunch of very useful information on jamming and granular media can be found in Aleksandar Donev PhD thesis (2006).
\end{itemize}

\subsection{Properties}
\begin{itemize}
 \item Spontaneously formed local nematic order.
 \item Size-, and position- depentend friction leads to streaming instability: fast cell remain small and escape trap, and slow cells grow and slow down even more.
 \item Cell bulcking: Cells grow linealry, and increase longitudal component of stress tensor(Anisotropic stress). 
 Then lateral displacement relieves compression.
\end{itemize}

\subsection{Contact Mechanics}
\begin{itemize}
 \item Central aspects:
 \begin{enumerate}
  \item Pressure and adhesion of acting perpendicular to the contacting bodies' surfaces(normal direction)
  \item Frictional stresses acting tangentially between the surfaces.
 \end{enumerate}

 \item Contact Mechanical Models:
 \begin{enumerate}
  \item Hertz: fully elastic model
  \item Bradley: purely Van der Waals model with rigid spheres
  \item JKR (Johnson, Kendall, Roberts): fully elastic model considering adhesion in the contact zone  - The JKR model considers the effect of contact pressure and adhesion only inside the area of contact. 
  Moreover the model includes histeresis effects - when pulled away they(objects like grains, cells etc.) elongate and stick beyond the distance that corresponds to the sum of their radii.
  \item DMT(Derjaguin-Muller-Toporov): fully elastic, adhesive andVan der Waals model -  contact profile remains the same as in Hertzian contact but with additional attractive interactions outside the area of contact.
 \end{enumerate}

 \item Non-adhesive elastic contacts - normal forces exerted by the surface mainly depend on:
 \begin{enumerate}
  \item depth of indentation(d)
  \item radius of the contact (a)
  \item effective elastic modulus: $\frac{1}{E^{*}} = \frac{1-\nu_1^2}{E_1}+\frac{1-\nu_2^2}{E_2}$, where $E_1,E_2$ are the elastic moduli,
  and $\nu_1,\nu_2$, are Poisson's ratios:
  $\nu = \frac{\mathrm{transverse ~ fractional ~ contraction}}{\mathrm{longitudinal ~ fractional ~ stretch}}$
 \end{enumerate}
 \item The assuption to the Hertzian contact theory:
 \begin{enumerate}
  \item The strains are small and within the elastic limit,
  \item Each body can be considered an elastic half-space, i.e., the area of contact is much smaller than the characteristic radius of the body,
  \item The surfaces are continuous and non-conforming, and
  \item The surfaces are frictionless.
 \end{enumerate}
 When these assumptions are violated, then it leads to problem called non-Hertzian. 
 However for large loads Hertz theory works moslty fine.
 
 \item Adhesive forces introduce changes to Hertzian model:
 \begin{enumerate}
  \item The area of contact was larger than that predicted by Hertz theory,
  \item The area of contact had a non-zero value even when the load was removed, and
  \item There was strong adhesion if the contacting surfaces were clean and dry.

 \end{enumerate}



\end{itemize}



\section{Ideas}
\subsection{Experiments}
\begin{enumerate}
 \item Maybe it is possible to inject GFP gene into yeast cell so it will shine when under mechanical stress - high pressure. 
 Shining may be triggered by some sort of (heat) shock protein.
 \item Lagrangian formalism is NOT well suited for systems where dissipative forces (e.g. friction) are present.
 \item Use the force: membrane tension as an organizer of cell shape and motility \cite{weiner:12} 
 Curvature-sensing proteins - they can sense the curvature of a cell membrane. 
 Curvature is a funciton of tension - higher tension, more energy to deform. Tagging with GFP/RFP may be usefull in tracing tensions in cells colony.
\end{enumerate}

%\subsection{Simulations based on Scherck's papers}
%\begin{enumerate}
% \item Simulate growing balls with some sort of repulsion: $V = \frac{K}{2}(r-\sigma)^{\alpha}$, where $\alpha=2.5$ is a Hertzian interaction.
% The reasoning behind $\alpha >2.0$ is that when two balls collide and deform one another, 
% the  interface between them, grows faster then the distance separating then gets smaller.
% \item $\alpha=2,\frac{5}{2},0$ - harmonic, Hertzian(in 3D), and hard-spehre repulsion.
% \item Simulate the box with a periodic boundary conditions. However if you want to introduce walls into your box
% do it by the interactions with mirror images (in the respect of the position of the imaginary wall). Be carefull
% around simulating the hole where balls can escape. 
% \item Seed the box with balls' seeds, and then grow them to the desired density.
% Monitor energy of the system. Calculate forces, and then if available relax the system: grow at the beginning and bisecting 
% \item Use different size balls, let's say the radius ration $r_1/r_2 \approx 1.4$
% \item Do attractive completely disrubt the scalling scenerio for the relaxation times that appear near point J for repulsive interactions.
 
%\end{enumerate}

\section{Papers}
\subsection{Jamming is not just cool any more\cite{nagel:98}}
\begin{enumerate}
 \item Jammed things: traffic, grainsand, powder raw material.
 \item vibrations from the pounding can reinitiate flow.
 \item When jammed, the disordered system is caught in a small region of phase space with no possibility to escape
 \item Jammed systems are fundamentally different. If the applied stress changes the structure by even a little, the whole jam breaks up. 
 \item One example of a system that is jammed and yet not fragile is foam.
 \item supercooled water, which freezes into glassy state can also be considered as jammed system - sampling noneq. limited space.
 \item Jamming can occure when density is high enough. 
 \item One then can unjam the system by rising the temperature(like in the case of supercooled water), or by applying stress. 
 \item ``Temperature`` - any kind of vibration (mechanical or thermal).
 \item Glass may have a lower glass transition temperature under high shear stress. 
 Likewise, a jammed granular material or foam may have a lower yield stress when random motions (that is, thermal fluctuations) are present.
 \item Conceptual framework from which to investigate the jamming transition, 
 suggesting a possible analogy between temperature and shear stress, the letter being considered alike a thermodynamic parameter.
 \item Examples of three axes:
 \begin{itemize}
  \item A liquid with low viscosity solidifies into a glass when temperature is lowered.
  \item Flowing foam becomes rigid when the applied stress is lowered.
  \item Colloidal suspension loses the ability to flow when the density is increased. 
 \end{itemize}
 \item The concept of jamming has been extended by \cite{weitz01}:
 \begin{enumerate}
  \item Attractive interactions might have the same effect in jamming the system as a confning pressure,
  supporting the concept of a jamming phase diagram for attractive colloidal particles, 
  furhter providing a unifying link between the glass transition, gelation and aggregation.
 \end{enumerate}


\end{enumerate}

\subsection{Highly evolved grains \cite{ohern:13}}
\begin{enumerate}
 \item They are far-from-equilibrium systems, as they are too large to experience thermal fluctuations and thus must be externally driven to induce particle motion.
 \item Packings of grains with wishbone shapes stiffen following a concave-up stress-strain curve, 
 and possess failure stresses that are more than a factor of three larger than those for spherical grains.
 \item Other relevant properties: density, shear modulus.Can be optimized as a function of interparticle forces, and packing protocol.
 \item Long-standing open task: identifying the most dilute jammed packings of spherical particles.
\end{enumerate}

\subsection{Repulsive Contact Interactions Make Jammed Particulate Systems Inherently Nonharmonic\cite{schreck:11}}
\begin{enumerate}
 \item When isostatic(in which the number of contacts is the minimum required for mechanical stability) systems lose even a single contact, they become fluidized.
 \item One-side repulsive potential make isostatic jammed system nonharmonic even in the limit of vanishing perturbation.
 \item Isostatic jammed system becomes nonharmonic when even single conact is borke. It occurs above the amplitude $\delta_c$, which average over all exctiation modes $\left<\delta_c\right>$ tends to zero with the number of particles
 \item Interactions potential: $V(r_{ij}) = \frac{\epsilon}{2} \left( 1 - \frac{r_{ij}}{\sigma_{ij}} \right)^2 \Theta\left( 1 - \frac{r_{ij}}{\sigma_{ij}} \right)$, where: $\sigma_{ij} = (\sigma_i + \sigma_j)/2$
 \item The MS(mechanically stable) packings were generated using the compression and energy minimization protocol.
 \item In nonharmonic regime the density of vibrational modes cannot be described using the dynamical matrix ($U_{x,y} = \partial^2 V / \partial x\partial y$)
\end{enumerate}

\subsection{Recent results on the jamming phase diagram \cite{con:10}}
\begin{enumerate}
 \item Soft, frictionless spheres, and jamming defined as a divergence in relaxation time.
 \item At zero temperature and zero shear stress (on the voluem fraction axis i.e. no particle motions), jamming is identified as appearance of mechanical strenght.
 \item The system accuaries mechanical strenth at J-point.
 \item Jamming transition at J-point has mixed first-order-second-order behaviour:
 coexistence of quantities varying continuously at the transition, such as the pressure or the shear modulus, 
 and of quantities changing discontinuously, such as the mean contact number per particle. 
 \item Jamming surface may be also considered as a surface of constant $\tau_g$, where $\tau_g$ is the longest available
 relaxation time to an experiment or simulation - but there is a debate on the shape at $\tau=\infty$.
 \item Some theoretical work suggest that true jamming for soft spheres system can occur only at T=0 (relaxation time diverge only at T=0).
 \item Fluc's in athermal system occur only when system is flowing. When system is jammed no fluc's can ever occur. This is in contrast with the
thermal case, in which fluctuations always occur, even if large fluctuations able to unjam the system may be so rare that they are not observed within any reasonable observation time.
 \item Jamming may occur not at the specific singular poing $\phi$, but in the range $\phi=0.636-0.646$. The reason is that jammed system is
 out-of-equilibrium so the behaviour may depend on the procedure. Thus jamming may occur in narrow, but still segmend of $\phi$.
 \item At $\phi=0.646$ which can be obtained by slow 'equilibrium' compression, because at thaat density pressure diverge.
\end{enumerate}

\subsection{Jamming by shear\cite{bi:11}}
\begin{enumerate}
 \item A jammed system can resist small stresses without deforming irreversibly.
 \item Jamming of frictional, disk-shaped grains can be induced by the application of shear 
 stress at densities lower than the critical value, at which isotropic (shear-free) jamming occurs.
 \item These jammed states are fragile when applied small shear strees - forces net percolate only in one direction.
 \item When applied certain minimum  shear stress forces netwrok percolates in all directions and system becomes jammed -this transition is controlled 
 by the fracion of force bearing particles which is independent on the density.
 \item On $\phi$ axis in generalized Liu-Nagel diagram there exist two values $\phi_J$ and $\phi_S$. The former is the value at which isotropic jamming occur,
 and below the latter no shear-jammin can occur. In between shear-jamming can occur. 
 \item The value of $\phi_J$ depends on the protocol of jammed system preparation. 
 \item Friction only slightly change the $\phi_J$, however it change number of contacts at jamming, more strongly than expected.
 \item Fragile states - they have a strong force network that carries the majority of the shear stress (deviatoric stress) 
 but which spans the system only in the compressive direction. 
 \item Sshear jammed states - are characterized by two non-parallel populations of force chains.
 \item The fraction of non-rattler grains ($f_{NR}$) controls the percolation of the force networks and emerges as the single parameter distinguishing between unjammed, fragile and shear-jammed states.
 \item Starting from an unjammed state, the strong force network undergoes two sequential percolation transitions controlled by the non-rattler fraction, $f_{NR}$
 \item Contact anisotropy(or stress anisotropy) acts as an order parameter. For values $\psi_J \gg \psi $it does not depend on $\psi$ but if goes to zero at $\psi_J$,
 thus at $\psi=\psi_J$ jammed states become isotropic.
\end{enumerate}


\subsection{The Jamming Transition and the Marginally Jammed Solid\cite{liunagel:10}}
\begin{enumerate}
 \item The simplest possible jamming model: frinctoinless spheres intercationg via repulsive finite-range forces at zero temperature.
 \item The simplest possible model:
 \begin{itemize}
  \item Spheres interact with a potential which vanish at some distance that defines their diamter
  \item At zero temperature, the system is always in mechanical equilibrium
  \item Jammed particles: any infinitesimal force will be resisted by the force network
  \item At a critical packing fraction the system is in between liquid and solid.
  \item The sharp transition exhibits discontinouity as in first-order phase transition, and power-law scaling , as in an ordinary critacal point. 
  \item Zero-temperature transition with aspects of both first- and second- ordeer behaviour, and multiple diverging and vanishing length scales. 
  \item The order parameter that characterize transition is Z, the average number of overlaps a particle has with neighbors. 
  This value jumps discontinuously,,to $Z_c$ at T=0 - because the be in place in mech. eq. it must be surounded by other particles from other sides. 
  \item For frictionless particles $Z_c=2d$. where $d$ is dimensionality.
  \item Below the cutting length: $ \mathrm{l}^{\star}\sim \Delta\phi^{-1/2}$ system looks isostatic, but above it, it's overcoordinated and should behave as a normal elastic solid - can be observed by probing responce to a point force.
  \item Pair-correlation function g(r) diverges at $\phi_J$ with first peak at $r=\sigma$ and amplitude $Z_c$, 
  which aslo diverges as $g(r) \sim \Delta \phi^{-1}$, and $g(r) \propto \sqrt{r-\sigma}$
  \item Numerically calculated exponents are the same in two and three dimensions indicate tahat the transition os mean-field-like. 
  Upper critical dimension of the jamming transition should be d=2.
  \item Each mecahnically stable configuration corresponds to a local minimum in the potential-energy landscape.
  \item Different minima or mechanically stable configurations can have different jamming thresholds, $\phi_C$. 
  The width of distribution vanish in the infinite system size. For 2D disks $\phi_J\approx0.84$ and for 3D spheres $\phi_J \approx 0.64$.
  However, different protocol can provide yield different results, since even in infinity there is still distribution of jamming threashold. 
  \item Jammed spheres at T=0 violate Debey law ($D(\omega)\sim\omega^{d-1}$). Instead $D(\omega)$ is plateau down to zero freq.
  \item The existence of low-freq quasi-localized modes suggests that stability of jammed solids may be different from that of ordinary crystals.
  \item a) Thermal excitations of the same energy gives larger amplitude (for some atoms) for these localized modes - thus making opportunity for breaking contacts.
  \item b) Moreover these low-freq modes have realatively small en. barier to overcome when they are deforming
  \item c) Finally low-participation ratio (mosty low-freq and high-freq modes) are highly unstable to compression.
  \item Different mechanically stable packings created with different initial conditions have the same properties: elastic moduli, coord. number, and other quantities depending on $\Delta \phi$.
 \end{itemize}
 \item Isostatic jammed states can be cleanly studied at T=0, because they are mechanically stable, at configuration which correspond to metastable minimum.
 \item Working at T=0 is appropriate for granular systems and foams, for which energy of even small rearrangement is many order of magnitude greater than thermal flucs.
 \item 
 \item Zero-temperature solid lose its rigidity by creating at least one soft-mode (i.e. mode with zero frequency). 
 In this case the failure of creates another solid (and not a fluid). The complete loss of rigidity should be reflected in behavior
 of just a single mode. The low-freq modes should display characteristic signature of jamming/unjamming. 
 \item Low-freq modes are dominant exctitaions, thus they also control how it responds to a small increase in temperature or applied stress - on top of a decrease in density.
 \item At temperatures greter than zero, a vanishing lenght scale - overlap distance - scales as: $\mathrm{l}_{w} \sim \phi - \phi_{\nu}(T)$, which is consisten with scaling at T=0.
 \item Once rotational degrees of freedom are introduced, the isostatic number in three dimensions increase from $Z_{iso}=6$ to $Z_{iso}=10$.
 However, at the jamming transition $Z_c$ does not jump discontinuously from 6 tp 10, when infinitesimally perturabtion to the shape is introduced. 
 In fact the number of interacting neighbors at the transition increases continously as the shape is varied.
 \item The resolution to the above is that soft modes (freq=0) localized at each sphere, are recruited in non-zero freq. band when deformation is applied.
 At small distortions, $\epsilon$, the resulting rot. modes for a new band that lies below the band of translational modes found for spheres. 
 Then, the onset freq. of the upper bound of modes scales as $\omega^{\star}\sim Z_c(\epsilon) -6$ - exctly the same scaling, that was observed forcompression. 
 \item Frictional spheres:
 \begin{itemize}
  \item Particles have in addition to normal forces $f_n$ also tangential forces $f_t$, which are up to a threashold set by the friction law: $f_t < \mu f_n$.
  \item The presence of tangential forces introduce another equations, thus change counting degrees of freedom.
  \item At the jamming theshold, meechanically stable packings can exist over a range of values: $d+1 \le Z_x \le 2d$.
  \item Contraty to frictionless particles, some properties of the system are not the same - they depend on the way the system was prepared!
  \item However, it is the case that static shear modulus obeys the expected scaling fot J point with $Z_c$ replaced by $Z_{iso}^{\mu}$.
  \item However, packing prepared genlty(and approaching the lowest density that can be typically be accesed for a given $\mu$) 
  tend to have the properties similar to those of frictionless particles sphere packings near the jamming threshold.
 \end{itemize}
 \item  Boson peak - it seems that it emerges as a result of appearance of new exctitations upon jamming.
 One of the property of boson peak exctitations is that these exctitation are poor transporters of energy.
 \item The ratio $C(T)/T^3$ of the heat capacity to the expected $T^3$ dependence (predicted by Debye model for xtalline solids), 
 exhibits a characteristic peak - known as boson peak.
 \item Anharmonic effects in jammed spheres packings are quite different from those in crystals. 
 \item Moreover, the low-frequency(and quasi-localized) modes of jammed systems are, considerably more anharmonic than the high-frequency ones. 
\end{enumerate}

\subsection{Computational Modeling of Synthetic Microbial Biofilms \cite{rudge:12}}
\begin{enumerate}
 \item A key factor in the efficiency and robustness of biofilms lies in their spatial organization. 
 \item For a typical colony there can be $10^4-10^5$ cells.
 \item Rigid-body method that includes growth of cells.
 \item In simulations of biofilms, each cell is coupled to many others through biophysical interactions and signaling. 
 \item Since growth occurs on a longer timescale tan biochemical interactions, growth is updated in discrete steps, and intracellular int. and signaling is solved separately.
 \item After each biophysical step, the state is of each cell is updated, and the intracellular and signaling systems are integrated forward by the appropriate time step.
 \item Rod-shaped bacteria, the shape approxitamed by cylider capped with hemispherical ends - capsule.
 \item In TYPICAL growth conditions cells exhibit very little deformation - thus cell lenght can be included as a degree of freedom (and the whole cell keeps the same shape).
 \item A cell is described as $\bar{x}(t) = (c_x, c_y, c_z, \phi_x, \phi_y, \phi_z, L)^T$, where $(c_x, c_y, c_z)$ is cell position, and $(\phi_x, \phi_y, \phi)$ cell orientation.
 Moreover each cell grows at a rate $\dot{L}$ - for small time periods growth is linearized -i.e. constant.
 \item In the low Reynolds number regime appropriate for bacteria, viscous drag dominates inertia - and cells move by distance proportial to applied impulse. 
 \item Upon cell division, a small amount of noise is added to the direction vecotr of each daughter cell - in account of imperfections in cell shape and alignment.
 \item Approximatio of experimentally observed variation in cells' lenght a division threashold is uniformly distribute 3.5-4.5 $\mu m$.
 \item Complex regulatory mechanics that determine cell behavious are sufficiently modeled based on empirical rules. 
 \item Each cell contains set of variables describing its state: position, direction, lenght, radius, volume, area, species, signals, celltype, growthrate, divideFlag. 
 \item Cell signaling is a key part of multicellular organization.  Signaling is introduced to th esimualtion by "medium".
 \item Difficulties in simulating large bacterial populations are: (i) numerical stability of the solution to the simulated system (can be solved with rigid-body dynamics)
 (ii) speed of simulation, which can be solved with e.g. - OpenCl and GPU.
\end{enumerate}

\subsection{Diffusion-Limited Growth in Bacterial Colony Formation \cite{MH:90}}
\begin{enumerate}
 \item When biological grotwh is governed by diffusion-limited processes, growing patterns show features such as screening, repulsion etc.
 \item Growth of bacterial(Bacillus subtilis) colonies on agar plates is governed by DLA processes contained in plates.
 \item Bacterial colonies growth is affected by environmental physical conditions and the way bacteria prolifirate. 
 \item On plates, with nutritients concentration 1g/L colonies develop in fractal structure with a dimension $\approx 1.72$ - obtained from box counting method.
 \item Branches screening effects observed - characteristic to pattern formation in Laplacian field.
 \item Two colonies spearated by small distance repeal each other, and neve fuse - characteristic for Laplacian field. 
 \item The Laplacian field is made by diffusin nutritients and not by some kind of growth-blocking metabolic waste. 
 \item In the limit of low nutrient concentrations the growth of bacterial colonies is generally controlled by diffusion-limited processes.
 \item DLA consists generally from two mechanisms: (i) Laplacian field and (ii) local growth mechanics. In case of bacterial colonies these are:
 (i) nutritien concentration and (ii) bacterial proliferation. Moreover, for too high nutrients concentration DLA pattern is not fomred -  diff. is not limiting factor any longer. 
\end{enumerate}

\subsection{Vibrations of jammed disk packings with Hertzian interactions \cite{schreck:13}}
\begin{enumerate}
 \item Dry granular media - discrete grains, interact via purely repulsive, frictional contact interactions.
 \item Sources of nonlinearity in dry materials include:
 \begin{itemize}
  \item the nonlinear form ofHertzian interactions betwen grainsans
  \item conact breaking and formation
  \item dissipation
  \item rolling and sliding frictional contacts
 \end{itemize}
 \item This manuscript concerns nonlinearitis that arise from the shape of Hertzian interactoins.
 \item (i) The shape of Hertzian interaction give rise to nonlinearities in the vib. response of jammed disk packings.
 Right after the first contact is broken the system creates set of discrete harmonics and beats among these and normal mode freqs, 
 unitl the moment where on average 1 conact is borken - then creates a continous spectrum of vibrations. 
 \item (ii) These nonlinearities are weak compared to those generated by contact breaking.

\end{enumerate}

\subsection{Localized excitations in a vertically vibrated granular layer \cite{swinney:96}}
\begin{enumerate}
 \item Oscillons - stable,localized 2D exctiations.
 \item Granular standing waves is a cooperative behaviour resulting from coolisions between grains. 
 Those waves can from stripes, squares, hexagonals etc. 
 \item Oscillons emerge under the same conditions as standing waves.
 \item An oscillon is a small circullar simetric excitation w/ a freq f/2 (f - freq of shaking the plate on which grains are placed).
 \item Oscillons are long lived objects. Moreover, since they are subharmonic craters and peaks can coexsist. 
 \item Oscillons do not propagate, but they drift slowly - they need about 1000 cycles to drift the distance of one diameter. 
 \item Oscillons of the same phase show short-range repulsions, and of the opposite phase-short range attration. 
 \item With the above "interactions" when they collide, they create stable structures: diploe, polumer, trimer, ion-lattice - Coordination numbers higher than 3 are not creater - for an isolated structre.
 \item Plannar patters can be considered in some cases as collection of oscillons. 
 \item Oscillons form in narrow range of frequencies, when dissipation and hysteresis are large. 
\end{enumerate}


\subsection{A cell-based simulation software for multi-cellular systems \cite{dd:10}}
\begin{enumerate}
 \item Multi-cellular systems are of multi-scale nature, and the research is shifting towards studies of whole cells or populations of cells
 \item Several milions of cells (in 3D) are simulated by  agent-based models.
 \item Each cell is modeled as an isotropic, elastic, and adhesive sphere.
 \item Each cell is capable of migration, growth and division.
 \item Cell-cell and cell-matrix is modeled by JKR model (validated by Chu et. al, 2005).
 \item Cells are parametrized by: 
 \begin{enumerate}
  \item Young's modulus, 
  \item Poisson ratio
  \item the density of membrane receptors -  responsible for cell-cell and cell-substrate adhesion 
  \item its (intrinsic) radius
 \end{enumerate}
 \item Prolifireting cells double their sizes, and deform into dumb-bells shape, before splitting into two daughter cells (Drasdo'95).
 When the exerted pressure is $p_tot > p_0$ pressure then cell goes into rest phase G0. The orientation of the cell divisio is random.
 
 \item Cell migration is modeled by Langevin equations: summarizing JKR force, friction forces (c-c, c-matrix: parametrized by friction coeff.), and optionally random forces.
 \item Equation for diffusion and consumption of nutrients, growth factors, etc. 
 on a lattice are solved using the Euler forward method 
 (fulfilling the Courant-Friedrichs-Lewy condition i.e. time step must be smaller than the value providing numerical stability of the solution)
 \item Comparsions between JKR, Hertz and DMT models suggests that the  precise form of the interaction force may have no significant impact on the multi-cellular dynamics.
 \item In the absence of chemotactic signals, isolated cells in suspension or culture medium have been observed to perform an active random-walk-like movement(Gruler, 94; Glazier'96).
 \item Cells parameters: diameter, diff const, energy of single adhesive bond, cell fric. const, cycle time, max. cycle time variation, Young modulus, densitiy of adhesion molecules. 
\end{enumerate}

\subsection{Johnson-Kendall-Roberts Theory Applied to Living Cells \cite{pin:05}}
\begin{itemize}
 \item When contacting surfaces adhere weakly and deform a little, then DMT approach works.
 \item At higher adhesion and deformability, when surfaces are upon separating forces, 
 there is a finite, nonzero contact area at separationt - then JKR theory ives the relation between the pull off force $F_s$ and the adhesion energy $W_{adh}$
 \item When the cytoskeleton of the cells has a complete 3D structure that maintains a slightly deformable spherical shape, 
 JKR theory is applicable to relate the separation force to the adhesion energy. In this case, when cytoskeleton is responsible for
 cells' spherical shape, the cells do not behave like shells but like elastic spheres.
 \item Adhesion(i.e. attraction) can be induced for example by depletion effect cause by nonadsorbing, water-soluble polymers.
 \item Model was tested JKR theory based on ahdesion forces when pulling away two cells. 
 \item Elastic behavior is expected up to the time, when cytoskeleton cannot reshape itself upon the force load (plastic flow, which take up to several minutes).
 \item 
\end{itemize}

\subsection{Mechanically Driven Growth of Quasi-Two-Dimensional Microbial Colonies \cite{bw:13}}
\begin{enumerate}
 \item The strength of mechanical interactions determines the speed with which the colony expands in space, with diffusion of the nutrient playing a secondary role.
 \item Cells are modeled as growing spherocylinders with a variable length.
 \item The colony grows on a 2D flat surface with nutrient concentration c(x,y)( diffusion with a diff. coeff. D).
 \item Nutrients are consumed at a rate $kf(c) = k\frac{c}{c_{half}+c}$ per unit biomass density, with half-saturation constant $c_{half}$.
 \item Cells grow (by elongation) at a rate $v_gf(c)$. 
 \item The cells interact mechanically  by the Hertzian theory of elastic contact.
 \item the nutrient becomes depleted within the colony so that only cells in a thin layer at the front are growing.
 \item nonmotile microorganisms replicate and push each other away as they grow.
 \item Mechanical interactions are understood to be very important in such systems; 
 in particular, mechanical pressure has been hypothesized to strongly affect the growth and apoptosis rates of cells, 
 leading to an alternative form of growth limitation.
 \item The density profile close to the edge decays according to a power law towards the uncompressed cell density $\rho_0$. 
 It is in striking contrast to Fisher-Kolmogorov waves, which exhibit exponential density profiles in the wave tip.
\end{enumerate}

\subsection{Buckling instability in ordered bacterial colonies \cite{tsimring:11}}
\begin{enumerate}
 \item Complex spatio-temporal organization of bacterial colonies may depend on:
 shape, motility, membrane structure, hemotaxis, cell,cellcommunication, and direct mechanical contact among cells in dense colonies.
 \item In biofilms or colonies in cavities, long-range signaling may be of secondary importance. 
 \item in close proximity, such as in biofilms, bacteria usually lose their flagellae and become non-motile.
 \item In large colonies the perfect nematic order is never reached; 
 multiple domains with different orientations are constantly regenerated in the bulk of the colony.
 \item Growing colony of nematically ordered cells is prone to the buckling instability - destroying  perfec nematic order. 
 This instability is related to the anisotropy of the stress tensor in the ordered cell colony.
 \item Buckling instability occurs at the place where the density of cells and the pressure are greatest.
 \item In continuum model granilarity of cells has been neglected, and cells were modeled as continous fileds: density $\rho(\mathbf{r})$ and velocity $\mathbf{v}(\mathbf{r})$.
 \item When cells are densely packed, the 'cellular fluid' is incompressible.
 \item Discrete-elements simulation:
 \begin{itemize}
  \item Each cell is represented by a spherocylinder with fixed diameter d and variable length l.
  \item Cells grow exponentailly unitl they reach critical length, at which every cell is replaced by two daughter cells.
  \item The cricital lenght is assigned randomly at cell's birth -  to break artifical synchornisation of cells division.
  \item The normal and tangential (frictional) forces moving the cells are computed based on the overlap of virtual soft spheres centered at the nearest points on the  axes of interacting spherocylinders.
  \item It is expected that the buckling instability may strongly affect the structure of bacterial populations 
  in confined environments (such as surface-bound biofilms), if significant internal stresses develop there due to the cell growth.
  \item Buckling istability may also affect the tissue growth and structure in multicellular organisms 
 \end{itemize}

\end{enumerate}

\subsection{Particle-Based Multidimensional Multispecies Biofilm Model \cite{pic:04}}
\begin{enumerate}
 \item Processes ongoing in biofils:
 \begin{enumerate}
  \item growth and decay
  \item division and spreading
  \item substrate transport and reactions
  \item biomass detahment
  \item liquid flow past the bioflm
  \item attachment
 \end{enumerate}
 \item Model:
 \begin{enumerate}
  \item Reactor contains two compartments: bulk liquid(completely mixed), and biofilm.
  \item Biomass consists of active(up tp \#B types) and inert particles(that result of the decay of active particles).
  \item Particle may represent an individual cell or cluster of cells.
  \item Particles are hard-spheres.
  \item The growth of every particle depends on accessible nutrients.
  \item The total mass in spherical particles is limited by $M_{max}$, which is independent of particle's type.
  \item When max mass limit is reached, new daughter particle is created, which touches mother particle in a random position, 
  and part of mother's mass is redistributed among daughter particle.
  \item PBC are used, and particles are spreading as a result of exerted forces by other particles.
  \item Concentration field is solved numerically including reaction(nutr. consumption) component.
 \end{enumerate}
 \item Investigated properties:
 \begin{enumerate}
  \item spatial dist. of substrate concentration and fluxes
  \item biofilm structure
  \item total biofilm mass
  \item dynamics of biomass distr.
  \item dynamics of biofilm detachment
  \item dynamics of solute concentration
  \item steady-state concentration of biomass and solutes in the biofilm
 \end{enumerate}
 \item A perfect steady-state with respect to the spatial dist. of particles and solutes is not reached.
 It is mainly due to stochasticity in the model, which comes from:
 \begin{enumerate}
  \item The initial distribution of biomass particles on the surface
  \item random choice of direction for the placement of a daughter cell
  \item uneven cell division
 \end{enumerate}
\end{enumerate}

\subsection{A single-cell-based model of tumor growth in vitro: monolayers and spheroids \cite{dd:05}} 
\begin{enumerate}
 \item Types of models:
 \begin{enumerate}
  \item single-cell-based (including cellular-automata, and off-latice)
  \item continuum models
 \end{enumerate}

 \item The model:
 \begin{enumerate}
  \item Basic unit of the model is a signle cell. Each cell is an elatic, sticky particle of limited compressibility and deformability.
  It's also capable of active migration, growth and division.
  \item During the divisin phase(mitosis) the cell has dumb-bell shape.
  \item Attractive and repulsive interactions are modeled using JKR model.
  \item Cells perform RW (in an absence of chemotactic signals) - characterize by diffusion constant D
  \item Friction-dominated overdamped stochastic motion is modeled with Metropolis algorithm.
  \item Glucose is limiting nutritent for the model of tumor spheroids in suspention.
  Cells can grow only if the local glucese contentration is above a certain threshold, and die by necrosis bolow other conc. threshold.
  \item Glucose is loally consumed with the rate $\hat{\gamma}$ and is described by partial-differential equation describing reaction-diffusion kinetics.
  \item The influence of extracellular matrix on cells migration is neglected.
  \item Average cell cycle time is influenced by nutritiens, regulatory factors, and mechanical stress.
  \item Cell can grow only when if it's not extensively deformed (or compressed) - i.e. the growth of the cell may be controlled by the compression sensed by cell's cytoskeleton.
  \item Three scenerios are possible:
  \begin{enumerate}
   \item Cell grows when a deformation is removed.
   \item A cell that was subjected to the critical deformation longer then some threashold does not enter the cell cycle again.
   \item If the critical deformation lasted longer than some specified time the cell undergoes apoptosis.
  \end{enumerate}

  \item Lysis of dead celly is modeled as a very fast or very slow process by: rmoving or not removeing at all a dead cell.
 \end{enumerate}
 \item Division of nonboundary cells(in tumer monolayer cells) is repressed by the neighbor contacts.
 \item The highest proliferation actvity is close to the tumor boundary - where the local concentration of glucose is the highest, 
 while inside the tumor a necrotic core forms.
 \item If at the same time each cell is supplied with glucose then cells divide everywhere hence the tumor grows exponentially fast.
 \item Simulations suggest that tumor grows exponentailly fast, but later is followed by a linear expansion of the tumor diamteres - both in layer and spherodi case.
 \item It is suggested, that linear growth regime is characterized by a cell number increase confied to a surface layer. 
 A nutritient limitation in tumor spheroids affects slighlty the expansion velocity.
 \item In case cells are of the average size and in absence of apoptotic or nectoric core, the linear expansion regime is accompanied by power law $N \propto t^d$.
 \item Cells at tumor boundary have performed more cell divisions and accumulated more mutation than cells in the interior, or in cells of a tumor of equal 
 size 
 \item Above a certain glucose concentration the growth kinetic in tumor spherodis should become independent of the glucose conc. and be determined by biochemical interactions only.
 \item The extend of exponential growth depends on cessation of cell division conditions.
 \item Increase of cell mobility (lower viscosity etc.) extends exponential growth phase, and increase the growth velocity.
 \item In case of cell-cell adhesion absence, tumor initially grows exponentailly fast,  while the aggregate diameter grows as $R \propto \sqrt{t}$ and then corss into linear regime
\end{enumerate}


\subsection{Dissipative particle dynamics simulations for biological tissues: rheology and competition \cite{elgeti:11}}
\begin{enumerate}
 \item Mechanical stresses couple to the orientation of the mitotic spindle during cell division (Julicher'07).
 \item Pressure strongly influences the expression of some genes (Farge'08).
 \item Cell differentiation depends on the mechanical properties of the substrace - in particular th value of its elastic modulus.
 \item Nonlinear elastic description of tissue has been very succesful.
 \item At the early stage of development, many tissues should be considered as liquids (finite viscosity and isotropic surface tension).
 \item The model:
 \begin{enumerate}
  \item Cell must reach some specified size, in order to enter mitosis stage - when it's reached (i.e. the distance between two particles creating a cell - see below) then cell divides
  \item Each cell is modeled as two particles interacting via a repulsive growth force - a distance between this two particles represent cell's size.
  \item Apoptosis acts randomly, removing cells with a constant rate $k_a$.
  \item Forces are divided into attractive and repulsive.
  \item Equations of motion are integrated with Dissipative Particle dynamics (Nikunen'03).
  The advantage oevr Langevin dynamics is that DPD does not assume  the background dissipation to be the dominant mode of dissipation - e.g. dissipative forces may be transfered to neighboriing cells.
  \item MY NOTE: alternatives to DPD are: fluctuating lattice-Boltzmann method(Ladd), or Stochastic Rotation Dynamics (Kapral). 
  A good description of DPD is in "Understanding Molecular Simulations, 2nd Ed., Chpt:17"
  \item Cell division and apoptosis break momentum balance. In that case a small background dissipation is needed.
  \item Particles belonging to the same cell act with repulsive potetnail. 
  Volume exclusion is modeled with a repulvie forces $\sim r^{-5}$ up to some cutoff value, and adhesion is modeled as a linear force. 
 \end{enumerate}
 \item Adhesive forces are at the origin of surface tissue tension that cause tissue aggregation an d round up.
 \item Cell division and apoptosis lead to fluidizaion of the tissue. 
\end{enumerate}


\subsection{Mechanical Control of Cell flow in Multicellular Spheroids \cite{dm:13}}
\begin{itemize}
 \item In embrology, collective migration occurs during morphogenetic transformations, in the absence of any cell proliferation.
 \item Cell migration is observed in tumors: in response to chemical cues, cancer cells can escape the primary tumor, and invide adjacent tissues. 
 The process is either collective and single cell.
 \item In tumors and spheroids, gradients of oxygen, grow factors etc. lead to increase of cell proliferatoin at the periphery, and apoptosis in the center, which drives cells flow.
 \item Growt rate of multicellular spheroids is drastically reduced by an external mechanical stress (Capello'11, '12), and this effect saturates at 5000Pa
 \item Cell division, rather than cell death or density, is affected by the mechanical stress.
 \item In steady-state spheroids, in the absence of mechanical stress cells flow toward the center is observed. 
 This convective motion is induced by cell division. 
 \item In growin spheroids, in the absence of stress moving back and forth toward the center was obseerved.
 It suggests that there is convergend flow, and divergent flow as well, but under the mechanical stress divergent flow is impeded. 
 \item The transport model:
 \begin{enumerate}
  \item Particles(markers) flux is: $J = \nu \rho$, where $\nu $ is the cell velocity, and $\rho $ is markers density
  \item diffusion of markers is neglected (they don't leave prolifiratnig cells)
  \item Prolifireting cells diffusion contribute to the displacement as $\mathrm{d}/\mathrm{R}$, 
  where $\mathrm{d}$ is cell's diameter and $\mathrm{R}$ is spheroid radius.
  \item Chemotaxis is also neglected
  \item Transport equation is: $\partial_t \rho + \nabla \nu \rho = 0$
 \end{enumerate}
 \item Growth rate decreased by 30\% under mechanical stress.
 \item The mode, which is involving purely convection fluxes, is in very good agreement with the experiment, 
 what suggests that for this system chemotaxis and diffusion are indeed negligible.
\end{itemize}

\subsection{Modeling the impact of granular embedding media, and pulling versus pushing cells on growing cell clones \cite{dd:12}}
\begin{itemize}
 \item Growing tumor overcomes mechnical barrier by exerting mechanical stress on the host tissue (biomechanics) or secreting matrix-degrageting enzymes, that modify local environment (biochemical).
 \item Biomechaniacal interactions play an important role in the growth control of tissues and tumors. 
 \item Increase of the cell population diameter and proliferation pattern can be explained by biomechanical form of contact inhibition, conrolled by force threshold,
 a pressure threshold, or a deformation threshold - above which cells become quiscent.
 \item Cells have the capacity to change thier physical properties by intramolecular control processes.
 \item In growing tumor two types of movement may exist:
 \begin{enumerate}
  \item pushing-type if cell division leads to local mechanical stress pushing neighbor cell in a direction of minimal mechanical compression
  \item pulling-type may be triggered by the emergence of a border
 \end{enumerate}
 \item Model:
 \begin{enumerate}
  \item Each cell is an isotropic, elastic and adhesive object.  
  \item Cells migarte, expand, grow and prolifirate.
  \item Cell interacts other cells and underlying substrate.
  \item Cell-cell interaction are described by JKR model (implicit equations in Parker'92) - however JKR neglects inhom. of cell material, 
  deviation from spherical shape, and plastic deformation which results from re-organizing processes of the cytoskeleton. 
  Also JKR model was used for adhesion forces between cells and underlying substrate (if present).
  \item Environment(in which living cells exsist) can be made of non-dividing cells or intert granular objects.
  \item Cells grow in steps $\Delta R$ unitl they double their size, and then divide.
  \item Dividing cell deforms into dumb-bell, and then two cells are created and separated. 
  \item If the total pressure exerted on cell is larger than a threshol value $p_0$, then it enters $G_0$ phase and do not divide. 
  When the pressure is less than the threashold cell again enters the cell cycle.
  \item A direction of dividing cell is random.
  \item Cells in suspenstion perform an active random-walk-like movement (Gruler'94).
  \item Cell migration is simulated as a friction-dominated over-damped motion with a stochastic contribution by a stochastic equation of motion for each cell.
  \item Two different types of motions are tested: unbiased(isotropic), and biased (in a direction of minimal stress - e.g. pressure).
  \item Despite the formal analogy to diffusion of colloids (Einsten relation). mechanism behid the motion is different, since cells regulate the motion on their own.
  \item For media, if objects represent not-dividing cells the dynamics is like in the case of living cells, and if objects are granular particles thn ethe active moving force is 0.
 \end{enumerate}
 \item Isolated monolayers show linear growth kinetics after an initial exponential growth phase. 
 \item Fingering patterns created by growing colony is quite robust in respoct to biomechanical properties of cells, and type of medium. 
 However changes to  biomechanical properties impact growth rate of the colony.
 \item Mecanical stress or strain-controlled cell cycle entrance mechanisms are able to explain many features of the growth patters 
 in monolayers and multi-cellular phenotypes - both in absence and presente viscous- and viso-elastic medium.
\end{itemize}

\subsection{Good random number generators are (not so) easy to find \cite{hell:98}}
"Anyone who considers arithmetical methods of producing random digits is, of course, in a state of sin."
--John Von Neumann (1951), Applied Math Series 12 (1): 36-38.
\begin{itemize}
\item From a practitioner's point of view, random number generators are good if they yiled the correct results in as many applications as possible.
\item Random number generatorsa are nothing more than deterministic algortihms that produce numbers with certain distribution properties.
\item RNGs task is to give correct results in tests which are relevant to user - they are not to simulate randomness. 
\item Two facts:
\begin{enumerate}
 \item With RNGs there are no guarantees, only predictions - it's because of finite sample size - indepentend of RNGs.
 \item Althogh there are no guaratnees, there are mathematical safety-measures against wrong simulation results. 
\end{enumerate}
\item Empirical testing - we can run application-specific tests with known theoretical results and then compare them to empirical results - obtained from simulation.
\item Theoretical support for RNGs: period length, structural properties, correlation behaviour etc. 
\item RNGs:
\begin{enumerate}
 \item linear congruential generator (LCG): $y_{n+1} = ay_{n}+b(mod ~ m)$, where m, a, b, and $y_0$ must be chosen.
 Random number in interval $x \in [0,1[$ is obtained by normalization: $x_n:y_n/m$.
 \item Multiple recursive congruental generator(MRG): $y_{n+k} = \sum_{j=0}^{k-1}a_j y_{n+j} (mod ~ m)$
 \item combined MRGs (cMRG): $x_n = x_n^{(1)} + x_{n}^{(2)} (mod ~ 1)$. $x_n^{(1)}$ and $x_{n}^{(2)}$ are already normalized values.
 \item twisted generalized feedback shift-register (tGFSR): $x_{n+p} = x_{n+q}\oplus x_{n}A$, $n \ge 0$,
 where w,p,q are parametres and A is $w\times w$ matrix with binary entries. This generator produce a sequence of w-bit integers.
 \item Inversive congruential generators (ICG): $y_{n+1} = a\bar{y}_{n} + b (mod ~ m)$, where $\bar{c} = 1/y$ or in other words $\hat{c} = c^{p-2} (mod ~ p)$.
 An important feature of ICGs is for prime modulus there is an absence of any lattice structure(for d-tuples).
 \item Explicit inverse congruential generaotrs (EICG): $y_{n} = \bar{a(n + n_0) + b} (mod ~ p)$.
\end{enumerate}
\item Theoretical support: period length, the intrinsic structure, and correlation analysis.
\item Random number generation is drawing without replacement, thus sample size should me much smaller than the period lenght of the generaotr.
\item In the case of linear generators square root is prudent upper bound.
\item Because of some intrinsic genrator structure, some types of simulation may fail.
\item Correlation analysis(CAs) allows reliable prediction of the performence of the samples produced by RNGs. 
CAs are nothing less than prototypes of simulation problems.
\item Parralel simulations create some additional problems. Basically there are following methods to generate RNs on parallel processors(more details in Anderson'90):
\begin{enumerate}
 \item Assign L different generators to L different processors
 \item Assign L different streams of one large-period generator to L processors
 \begin{enumerate}
  \item "leap-frog" - we assign the substream $(x_{nL+j})_{n\ge0}$ to processor j
  \item "Splitting" - we assign whole segment $(x_n)_{n\ge n_j}$ to processor j, where $n_1,...,n_L$ is a set of initial values to assure disjointness of the substream.
 \end{enumerate}
\end{enumerate}
\item Technique (i) cannot be recommended in general, since no resulsts on correlations between different random number generators(except EICG) have been reported.
\item "Leap-frog" may be unsafe since some RGNs subsequences (e.g. LCG($2^{48},55151000561141,0$)) perform even worse whan RANDU in spectral test.
Then if this subsecence is assigned to one of the processor, it may have serious consequences. 
\item Splitting is unsafe also in the respect of lurking long-range correlations.
\item RNG checklist
 \begin{enumerate}
  \item Theoretical support: period lenght, structural properties, correlation analysis.
  \item Empirical evidences: variable sample size, two-or higher level tests
  \item Practical aspects: tables of parameters available, portable implementations, parralelization techniques, large samples available
 \end{enumerate}
\item In general there is no safe generator. All have some unwanted sideeffects. 
\item Good RNG is characterized by theoretical support, convincing empirical evidences, and positive practical tests. Such RNGs produce correct results in many, though not all, simulations.
\item Three rules for using RNGs should be kept in mind:
\begin{enumerate}
 \item Don trust simulation results produced just by one RNG. Check the results with different RNG.
 \item Do not combine, vectorize, or parralize random number generators without theoretical and empirical support
 \item Get to know the properties of your random number generators.
\end{enumerate}

\end{itemize}

\subsection{Dissipative Particle Dynamics \cite{ums:02,war:95, war:97}}
\begin{itemize}
 \item To study hydrodynamics the times scales needed are much larger than water relaxation(or collision) time $\tau$, 
 and considered length-scales L are also much larger than the size of each solvent molecule(or interparticle distances). 
 Thus the idea is to drop MD since it also provides too much details, which are redundant in studying hydrodynamics.
 \item Molecuels are lumped into "fluid particles" in much the same spirit as the renormalisation group has been applied in polymer physics.
 \item Hydrodynamics means that transport equation for the momentum density field, is coupled to the transport of mass.
 \item On the other hand, the energy during the simulation is not conserved(is dissipated) what may be the problem when
 there is a temperature gradient in the system.
 \item Gallilean invariance impose $F_D$ and $F^R$ to be only dependent on: $r_{ij} = r_i - r_j$, and $v_{ij} = v_{i} - v_{j}$
 and further requirement of isotropy emands that the forces should transform under rotations as vectors.
 \item Finally, it is required that the dissipative forces $F^D$ are linear on the momentum and the 
 random forces $F^R$ are independent of the momentum.
 \item The total momentum conservation follows from symmetry property: $\zeta_{ij} = \zeta_{ji}$
 \item Random force $F^R \sim \Delta t^{-1/2}$ ! The dependence of random force on $\Delta t$ is due to 
 the agreement between MSD (wchich must be independent on $\Delta t$) - up to a friction factor - and number of time steps N (random kicks) applied to a particle - it can be calculated from the
 average mean square force transfered to the particle.
 \item There is no statistical difference  between the simulations using uniform random numbers and the simulations using Gaussian random 
 numbers. Since uniform random numbers take less CPU time to generate than Gaussian random numbers do, a choice for uniform noise should be made.
 
 \item In modyfied velocity-verlet model preditions for the veloicty in the future is made (by applying present forces with strenght $\lambda$),
 and correct for this in the final step.
 
 \item Brownian motion of uncharged colloids does not depend on the atomic details of the solvent.
 \item The only properties of the medium that matter are: temperature, density, viscosity.
 \item What we need is a model such that:
 \begin{enumerate}
  \item exhibits hydrodynamic properties.
  \item has thermal fluc's that can drive Brownian motion.
  \item is cheap to simulate.
 \end{enumerate}
 \item DPD reproduces correctly hydrodynamic behaviour - however there is no rigorous demonstration that this is true, 
 depsite the fact that there is large body of numerical studies suggesting that at $\delta t \rightarrow 0$ DPD fluid is described by the Navier-Stokes equation.
 \item To ensure, correct canonical distribution (in equil. simulation) dissipation and random noise are bounded by two relations: 
 $\omega^{2}_{r} = \omega_{D}(r)$ and $\sigma^2 = 2k_BT\gamma$. In the first condition, one of the function may be chosen arbitrarily, and second must follow from the relation.
 The second relation sets the relation between temperature and fluctuations.
 \item BD does not conserve momenta. In fact, the only thing BD conserves is number of particles, and thus only diffusive behavior can be observed.
 \item Time reveribility is also importnat in a DPD simulation, since only with a time-reversibility integration scheme can detailed balance be obeyed.
\end{itemize}

\subsection{Efficient numerical integrators for stochastic models \cite{cov:06a,cov:06b}}
\begin{itemize}
 \item A general method for deriving deterministic integratirs is based on the Trotter expansion.
 \item This expansion preserve the symplectic structure of the dynamics and conserve the dynamical invariants.
 \item Exact trotter integration requires generation of "coloured" noise.
\end{itemize}

\subsection{Measuring Cell Forces by a Photoelastic Method \cite{cur:07}}
\begin{itemize}
 \item Forces between groups of cells are likely to have complex and rather unpredictible distributions.
 \item External mechanical stimualtion stimulates the cell cycle [Curtis and Seehar'78]
\end{itemize}



\subsection{Dissipative-particle-dynamics model of biofilm growth\cite{scheibe:11}}
\begin{itemize}
 \item For biofilm growth two types of nutritents are required: electrons-donor and electrons-acceptor.
 Assuming that one of them is unlimited excess, the problem of food supply is reduced to only one substrate.
 \item The concentration of substrate is governed by advection-diffusion-reaction equation:
 \begin{equation}
  \frac{\partial C_s(\mathbf{x},t)}{\partial t} + \mathbf{V}(\mathbf{x},t) \cdot \nabla C_s(\mathbf{x},t) = D_s(\mathbf{x},t) \nabla^2 C_s(\mathbf{x},t) -r_s(\mathbf{x},t)
 \end{equation}
 where: $\mathbf{V}$ is liquid velocity, $C_s(\mathbf{x},t)$ is substrate concentration, $D_s$ is substrate diffucion coefficient (it may be uniform over all space or vary),
 and $r_s(\mathbf{x},t)$ is substrate consumption rate in biofilm phase. 

 \item Substrate consumption is given by Monod-type function:
 \begin{equation}
  r_s = \frac{\mu_m}{Y_{bs}} \frac{C_bC_s}{(K_s + C_s)}
 \end{equation}
 where $\mu_m$ is the maximum biomass growth rate, $K_s$ is the substrate saturation constant, and $Y_{bs}$ is the biofilm yield coefficrint. 

 \item The liquid velocity field is given by Navier-Stokes equations:
 \begin{equation}
  \nabla \cdot \mathbf{V} = 0
 \end{equation}
 \begin{equation}
  \frac{\partial \mathbf{V}}{\partial t} + \mathbf{V} \cdot \nabla \mathbf{V} = -\nabla P/\rho + \nu\nabla^2 \mathbf{V}
 \end{equation}
 \item The kinetic model of biofilm growth describing biofilm growth is given as:
 \begin{equation}
  dC_b/dt = Y_{bs}(r_s - m_sC_b)
 \end{equation}
 where $m_s$ is maintenance coefficient representing the biomass decay effect. 
 \item The biomass density has its maximum value $C_{bm}$, and whenever the biomass density $C_b$ grows larger than the threshold, the extra biomass
 is redistributed giving rise to biofilm volume expansion. 
 \item Liquid, biofilm and substratum(immobile background) are modeled as three different types of mesoparticles.
 \item Solutions to N-S equations(for $\mathbf{V}$ and $\nabla \mathbf{V}$) 
 are approximated by the low Much number flow of a slightly compressible liquid represented by DPD particles, and with velocity-verlet algorithm.
 \item each DPD particle carries specified mass of biofilm and substrate. They are specified by substrate concentration $c_{s,i}$ and 
 biomass density $c_{b,i}$. The change of substrate concentration is given by advection-reaction-diffusion equation.
 Thus, on top of consumption of substrate due to biofilm growth, there is alos dissipative and random exchange between neighboring DPD particles.
 \item Biomass particles are all for which $c_{b,i} >0$. When this value exceeds maximum biomass density $c_{bm}$ then that excess is instantenoulsy
 transferred to neighboring fluid particles, and then they become biomass particles as well. 
 (The exceess is assumed to be lost when there is no[withing a cutoff] liquid particle where biomass may be transferred).
 \item However, because biofilm is growing moslty at the interface liquid:biofilm in the course of simulation it is unlikely that
 biofilm particles may be discarded.
 \item In the above model , biofilm is coarse-grained itself, what means one DPD is a collection of many cells.
\end{itemize}

\subsection{Force distribution in a granular medium \cite{nagel2:98}}
\begin{itemize}
 \item Force chains and spatially inhomogeneous stress distributions are characteristic of granular materials.
 \item Spatial probability distribution P(F) for finding a normal force of magnitude F against a wall 
 decays exponentially for forces larger than the mean   $\bar{F}$ - what means that in comparison with Gaussian  distribution
 the probability of finding large forces is much higher. 
 \item For $F > \bar{F}$ distribution of forces decays exponentailly, whereass for $F < \bar{F}$ it flattens out and approach a constant value.,
 and increase toward force equal to 0.
 \item Neither boundary conditions posed by the vertical container walls 
 nor spatial correlations in the arrangement of beads due to crystallization near a wall during system preparation 
 influences the distributions of forces P(F).
 \item There is no correlations between laterla forces on different beads. 
 \item The exponential tail of the distribution can be understood on the basis of a scalar model (q model),
 where it emerges as a result of a randomization process that occurs as forces are transmitted through the bulk of the bead pack.
 \item However original q model predicts nonmonotonic behavior at $f\rightarrow0$ with $P(f)\rightarrow0$. 
 The behaviour remains even if the forces correlations in a plane corection is introduced.
\end{itemize}


\subsection{Contact force measurements and stress-induced anisotropy in granular materials\cite{bob:05}}
\begin{itemize}
 \item The force network carrying the external load is different for shearing and isotropic compression - and is history dependent due to friction.
 \item Forces resulting from these two loads have different angular (normal and tangential to the stress sdirection) distributions of contact forces,
 patial correlations depending on the macroscopic preparation.
 \item Forces have long-range correlations in the directtion of f-chains for shear loads, and are correlated over a short range for isotropic 
 compression regardless of a direction. 
 \item Normal forces distribution has peak around mean $\bar{F}$, exponential tail, and a dip for smaller forces (in contrast to Nagel'98 paper).
 \item For isotropically compressed systems, the normal force distribution dips towards zero for forces below the mean, 
 is broad around the mean, and decays  faster for large forces compared to the sheared system. 
 The tangential force distributions have a nearly exponential tail for forces larger than the mean 
 for both the sheared and the isotropically compressed system.
 \item Mean of tangential forces is an order of magnitude less than the mean for normal forces. 
 \item For both cases a distribution of Culomb failure criterion($S = |F_t|/\mu F_n$; $\mu$ is static friction coefficient)
 shows that most of the contacts is below Coulomb failure condition (i.e. it does not slip).
 \item Shear load presenst long range forces correlations  (up to 15 beads diameter) in the direction of applied force,
 and short range correlations in a perpendicular direction - correlations were calcualated for forces larger than mean i.e. in exp. tail.
 \item n complete contrast to sheared systems, the isotropically compressed system has nearly uniform force correlation in all directions.
 \item Sheared systems exhibit not only geometric (distribution of contacts) but also mechanical anisotropy (distribution of forces).
\end{itemize}

\subsection{Improving the Density of Jammed Disordered Packings Using Ellipsoids\cite{don:04}}
\begin{itemize}
 \item Packing problem - how densely objects can fill a volume (economically important problem of how much grain a barrel can hold).
 \item For equal spheres the solution is FCC packing with $\phi = \pi/\sqrt{18} \approx 0.74$
 \item Amorphous, jammed, random packings $\phi \approx 0.64$.
 \item Elispsoids can pack randomly more densely, up to $\phi \approx 0.64-0.71$, and in some cases(asoect rations) even up to 0.74.
 \item The problem with random close packing(RCP) is that a definition of what is random is lacking. 
 \item Some new concept - maximally random jammed (MRJ) state, corresponding to the least ordered among all jammed packings.
 \item However RCP and MRJ provides very similar results.
 \item For ellipsoids $\phi$ and $Z$- contact number - increases rapidly up to 0.71 and 10, respectiely, and some of them even up to 0.74.
 \item Despite of the very fast increase, packing density decreases again for very large or small aspect ratios - strong exclusion-volume 
 effects in orientationally disordered packing.
 \item This increase has no significatn packing ordering, defined as nematic parameter: $S_2 = \left< 2\cos^2 \theta - 1 \right>$.
 \item Packings are jammed according to the rigorous hierarchical definitions of local, collective, and strict jamming.
 These definitions are based on mechanical stability conditions that require that there be no feasible local or 
 collective particle displacements and/or boundary deformations.
 \item Packing density increases rapidnly as we deviate from shpere shape as a result of conact number increse - 
 due to additional rotational degrees of freedom of the ellipsoids.
 \item More contacts per particle are needed to eliminate all local and collective degrees of freedom and ensure jamming,
 and forming more contacts requires a denser packing of the particles.
 \item It's been conjectured that for frictionless random packings have just enough constraints to completely statically define the system, Z=2f,
 where f is the number of degrees of freedom (f=3 for spheres, 5 for spheroids, 6 for general ellipsoids and so forth).
 \item If friction is strong, then fewer contacts are needed, Z=f+1.
 \item Random packing shows  nonanalytic behavior as only aspect ratio deviates from 1. 
 However, crystal packings of spheres and ellipsoids show no such singular behavior.
\end{itemize}
 
 \subsection{Controlling active self-assembly through broken particle-shape symmetry \cite{dunk:14}} 
 \begin{itemize}
  \item Four types of particles considered: convex polar($s^+$) and antipolar($s^-$), and crescent-shaped($c^+$) and anticrescents($c^-$). 
  \item Polarity (fore-aft symmetry) defined based on relation between the shape and a direction of motion.
  \item $s^+$ particles align and do not change significanlty the direction and speed of motion - contarty to $s^-$ particles. 
  \item $s^+$ swarm and move cooperatively along the common axix.
  \item $s^-$ tend to form droplets that nucleate slowly from an initially homogeneous suspension.
  \item At very small or large $\phi$ differences are smaller since for low values partciles bahave as independent, and for larger
  densities polarities are too strongly hindered due to packing effects.
  \item Mixture of $s^+$ and $s^-$ (with the same speed) demixes and creates condensed droplets of $s^-$ and loosely moving $s^+$
  \item Created droples are almost completly devoided $s^+$.
  \item Droples as a whole are moving slowly in comparison to $s^+$, whereas particles in the moiddle of droplet move faster then particles on the edge.
  \item The morphology of the liquids depend on the characteristic shapes: $c^+$ and $c^-$. 
  For example $c^-$ create clockwise or counter-clockwise spinning rotors.
  \item Similariliy to mixtures of $s^+$ and $s^-$, when $c^+$ and $c^-$ are mixed they segregate into different phases. 
  \item $c^-$ rotors are stable against thermal fluctuations(noise), whereas the small c+ clusters decay rapidly in the presence of noise.
 \end{itemize}
 
 \subsection{Quantitative and qualitative microscale distribution of bacteria in soil \cite{rr01}}
 \begin{itemize}
  \item Soil suppots life of $10^9$ bacteria cells per gradient
  \item 4000-7000 different bacteria species 
  \item biomass 300-3000kg per ha.
  \item The cohesion between the mineral particles is ensured by organic cements such as microbial exopolysaccharides.
  \item Soil pores accounting for up to 50\% of the total soil volume.
  \item The volume of soil spaces depends more on the size, shape and stability of aggregates than on soil mineral particles themselves.
  \item Generally 40-70 \% of all bacterial are located in pores $2\mu m - 20 \mu m$.
  \item Distribution of different bacteria species is correlated with the amount of available water, its circulations, amount of oxygen, and availability of organic materials.
 \end{itemize}
 
 \subsection{Getting into a Proper Jam \cite{corwin12}}
 \begin{itemize}
  \item Commentary to van Hecke et al., PRL 2012 and Nagel et al., PRL 2012
  \item Properly jammed system must be stable to compression and to shear. 
  \item Previous compression-only algorithm suffer from shear instability due to neglecting shear by deformation of the boundaries of the volume - this deformation is an additional degree of freedomone whose energy has not been accounted for and minimized in the simulation.
  \item In d dimensions and N partilcles jamming means that every degree of freedom must be constrained. 
  For jammed systems these constraints are provided by contacts between particles and thus there should be $d \cdot N - d$  (-d since there is translational symmetry).
  However finite systems also need one extra constraint for positive bulk modulus and d(d+1)/2 - 1 constraints for shear modulus.
  \item Jamming is hypothesized to be  either a random first-order or a mean-field second-order transition - since bulk and shear moduli 
  scale (above jamming point) with the excess number of contacts between particles and colapse onto the same curve for different system sizes.

 \end{itemize}

 \subsection{Two regimes, maybe three? \cite{ds06}}
 \begin{itemize}
  \item The mechanical responses of cells are determined by the intrinsic mechanical properties of the cytoskeleton(CSK) material.
  \item Cell behaves like an actin gel, but only at small timescales.
  \item Powerlaw relationship has been found for cell deformation upon constant or oscillatory  mechanical stress. 
  \item Dynamic modulus of the cell scales with the frequency of the applied stress following two distinct regimes: actin network dynamics and soft-glass rheology(SGR).
  \item At high frequency ($>$100 Hz) - consistent with actin network dynamics, a universal power-law exponent is $\alpha=0.75$.
  \item At low frequency ($<$100 Hz) - SGR -  the modulus scales with frequency with a much smaller exponent $\alpha=0.05$.
  \item  This suggests that the fast dynamics of actin networks is of little relevance because its influence is important only over 
  short timescales ($<$0.01 s), whereas in cells most integrated mechanical events (spreading, crawling, contracting) are set within much longer timescales.
  \item One question remains: if SGR dynamics extends for $>100$s? It is motivated by earlier studies where slow SGR dynamics increases sharply between 10s-100s.
  \item Powerlaw exponent $\alpha$ decreases for increasing prestress of CSK. $\alpha$ escribes deformability; i.e. $\alpha=0$ for elastic solids
  and $\alpha=1.0$ for liquids.
  \item The fact that $\alpha$ changes with prestress suggests mechanism of regulation between solid-liquid behavior, and the origin of this behavior 
  may be related to the architeture of the actin network.

 \end{itemize}
 
 \subsection{The Syncytial Drosophila Embryo as a Mechanically Excitable Medium \cite{liu:13}}
 \begin{itemize}
  \item The early embryos of many species, exhibit metachronous mitosis, in which mitosis progresses as a wavefront through the embryo.
  \item Mitotic wavefronts are observed in cycles 9 through 13, and cycles of nuclei are strongly coupled cross the entire embryo by mitotic wavefronts.
  \item Metaphase and anaphase wavefronts exhibit identical wavefront patterns.
  \item Syncytial Drosophila embryos may constitute an example of mechanical signaling via nonlinear wavefronts that could also arise in very different biological contexts.
  \item The main idea/hypothesis is that: mechanical wavefront triggers mitosis. This mechanical wavefront is due to mitosis methaphase/anaphase which mechanically  affects neighbor cells.  
 \end{itemize}

 \subsection{Cell shape and cell-wall organization in Gram-negative bacteria\cite{w08, w11a, w11b}}
 \begin{itemize}
  \item The cell wall is both necessary and sufficient for determining cell shape.
  \item Cell wall is a collection of nodes conntected by springs - it is represented as a mesh with a void interior.
  \item There are three types of forces in the system: spring extension forces, bending forces, osmotic force.
  \item Springs extension and bending is calculated as usual - i.e. similar to MD simulations.
  \item Osmotic force is defined for the ith node as: $\mathbf{F}_{i,\Pi} = \Pi \nabla_i V$.
  \item To calculate the forces due to the osmotic pressure differential across the cytoplasmic membrane, 
  surface mesh first divide each hexagon of the spring network into six triangles defined by the edges of the hexagon and its center of mass. 
  Then the volume V of the cylinder can be calcualated by summing over the volumes of the prisms defined by the triangles on 
  the surface and the center of the cylinder.
 \end{itemize}
 
 \subsection{Collective migration and cell jamming\cite{fjj13}}
 \begin{itemize}
  \item Plithotaxis is the tendency for each individual cell within a monolayer to migrate along the local orientation of the maximal principal stress,
  or equivalently, minimal intercellular shear stress.
  \item Plithotaxis requires force transmission across many cell-cell junctions and therefore is an emergent property of the cell group.
  \item When proteins responsible for cell-cell junctions are knocked-down the tendency to plithotaxis decreases.
  \item  Within the cell monolayer intercellular forces are heterogenous and are not tied to neither particular position nor cell type.
  \item Stress peaks emerge from cooperativity across clusters of roughl 10-50 cells.
  \item As cells density gets higher there are increasingly large regions over which cells have to move 
  in a cooperative manner, rearrangements within cell packs must become increasingly cooperative, bigger and slower as well as more intermittent.
 \end{itemize}




\thispagestyle{empty} % No slide header and footer

\bibliographystyle{unsrt}
\bibliography{article}

\clearpage

\end{document}

%-Cruavzte ssh --progress --no-whole-file --stats --sparse --exclude *.svn*

